---
title: "Getting started with TCGAretriever"
author: "Damiano Fantini"
date: "December 17, 2019"
output: html_document
vignette: >
  %\VignetteIndexEntry{Getting Started with TCGAretriever}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
  % \VignetteDepends{TCGAretriever}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(graphics)
library(utils)
```

*TCGAretriever* is an R library aimed at downloading genomic data from cBioPortal (<https://www.cbioportal.org/>), including The Cancer Genome Atlas (TCGA) data (free-tier data). TCGA is a program aimed at improving our understanding of Cancer Biology. Several TCGA Datasets are available online. TCGAretriever helps accessing and downloading TCGA data in R via the cBioPortal API. Features of TCGAretriever are: 

- it is simple and reliable 

- it is tailored for downloading large volumes of data

Below, you can find a brief tutorial that descibes a simple use case to help ou getting started with *TCGAretriever*.

#### Cancer Study IDs

```{r fig.align='center', fig.width=6, fig.height=5}
library(TCGAretriever)

# Obtain a list of cancer studies from cBio
all_studies <- get_cancer_studies()

# Find published TCGA datasets
keep <- grepl("tcga_pub$", all_studies[,1])
tcga_studies <- all_studies[keep, ]

# Show results
head(tcga_studies[, 1:2])
```

```{r fig.align='center', fig.width=6, fig.height=5}
# Define the cancer study id: brca_tcga_pub
my_csid <- "brca_tcga_pub"
```

#### Genetic Profiles (Assays) and Case Lists

```{r fig.align='center', fig.width=6, fig.height=5}
# Obtain genetic profiles
blca_pro <- get_genetic_profiles(csid = my_csid)
head(blca_pro[, 1:2], n = 8)
```

```{r fig.align='center', fig.width=6, fig.height=5}
# Obtain cases 
blca_cas <- get_case_lists(csid = my_csid)
head(blca_cas[, 1:2])
```

#### Retrieve Genomic Data

```{r fig.align='center', fig.width=6, fig.height=5}
# Define a set of genes of interest
q_genes <- c("TP53", "MDM2", "E2F1", "EZH2")
q_cases <- "brca_tcga_pub_complete"
rna_prf <- "brca_tcga_pub_mrna"
mut_prf <- "brca_tcga_pub_mutations"

# Download RNA
brca_RNA <- TCGAretriever::get_profile_data(case_id = q_cases, gprofile_id = rna_prf, glist = q_genes)
```

**NOTE**: the resuting data.frame includes ENTREZ_GENE_IDs and OFFICIAL_SMBOLs as first and second column.

```{r fig.align='center', fig.width=6, fig.height=5}
head(brca_RNA[, 1:5])
```

```{r fig.align='center', fig.width=6, fig.height=5}
# Set SYMBOLs as rownames
# Note that you may prefer to use the tibble package for this
rownames(brca_RNA) <- brca_RNA$COMMON
brca_RNA <- brca_RNA[, -c(1,2)]

# Download mutations (simple)
brca_MUT <- TCGAretriever::get_profile_data(case_id = q_cases, gprofile_id = mut_prf, glist = q_genes)
rownames(brca_MUT) <- brca_MUT$COMMON
brca_MUT <- brca_MUT[, -c(1,2)]

# Show results
brca_RNA[,1:6]
brca_MUT[,1:6]
```

**NOTE**: when using the same `case_list_id` to retrieve different types of data (genetic profiles) results have consistent structure. In other words, data.frames include info for the same list of cases (and hence, the resulting data.frames have the same number of columns, and identical column names).  

```{r fig.align='center', fig.width=6, fig.height=5}
# Note that the columns (cases) are identical 
# and have the same order in both data.frames
sum(colnames(brca_MUT) != colnames(brca_RNA))
```

#### Examples and visualizations

**Relationship between E2F1 and EZH2 in BRCA**

```{r fig.align='center', fig.width=5.5, fig.height=5.8}
# Coerce to data.frame with numeric features 
df <- data.frame(t(brca_RNA), stringsAsFactors = FALSE)
for(i in 1:ncol(df)) { df[, i] <- as.numeric(df[, i])}

# Visualize the correlation between EZH2 and E2F1
with(df, 
     plot(E2F1, EZH2, 
          pch = 19, cex = 0.5, main = "E2F1-EZH2 correlation in BRCA"))
```

**Relationship between MDM2 and P53, by P53 mutation status in BRCA** 

```{r fig.align='center', fig.width=10, fig.height=5}
# Coerce to data.frame with numeric features 
df <- data.frame(t(brca_RNA), stringsAsFactors = FALSE)
for(i in 1:ncol(df)) { df[, i] <- as.numeric(df[, i])}
df$TP53.status <- as.factor(ifelse(brca_MUT["TP53",] == "NaN", "WT", "MUT"))

# Split data based on TP53.status 
lst <- split(df, f = df$TP53.status)

# Visualize the correlation between MDM2 and TP53 by P53 mutation status
par(mfrow = c(1, 2))
for(x in names(lst)) {
  with(lst[[x]], 
       plot(TP53, MDM2, 
            pch = 19, cex = 0.5, 
            xlim = c(-2.6, 2.2), ylim = c(-1.6, 3.6),
            main = paste0("MDM2-vs-P53 in ", x , " P53 tumors")))
}
```

